---
title: "Assignment4_PML"
author: "Nandan Tavargeri"
date: "23/10/2021"
output: html_document
---

##Synopsis

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.

The goal of this project is to predict the manner in which they did the exercise. This is the `classe` variable in the training set.

## Data description

The outcome variable is `classe`, a factor variable with 5 levels. For this data set, participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in 5 different fashions:

- exactly according to the specification (Class A)
- throwing the elbows to the front (Class B)
- lifting the dumbbell only halfway (Class C)
- lowering the dumbbell only halfway (Class D)
- throwing the hips to the front (Class E)

## Initial configuration

First we import the required modules into the program. We follow this by loading the dataset into two seperate dataframes, namely the “training" and “testing” dataframes

```{r configuration, echo=TRUE}
#Importing Modules
library(caret)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(RGtk2)
library(rattle)
library(randomForest)
library(gbm)

#Loading the datasets
train_url <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test_url  <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
init_org_training_data <- read.csv(url(train_url))
init_org_testing_data <- read.csv(url(test_url))
dim(init_org_training_data)
dim(init_org_testing_data)

```

## Data processing
Here the data is cleaned

```{r dataprocessing, echo=TRUE}
non_zero_var <- nearZeroVar(init_org_training_data)
org_training_data <- init_org_training_data[,-non_zero_var]
org_testing_data <- init_org_testing_data[,-non_zero_var]
dim(org_training_data)
dim(org_testing_data)

na_val_col <- sapply(org_training_data, function(x) mean(is.na(x))) > 0.95
org_training_data <- org_training_data[,na_val_col == FALSE]
org_testing_data <- org_testing_data[,na_val_col == FALSE]
dim(org_training_data)
dim(org_testing_data)

org_training_data <- org_training_data[,8:59]
org_testing_data <- org_testing_data[,8:59]
dim(org_training_data)
dim(org_testing_data)

```

## Cross-validation
In this section cross-validation will be performed by splitting the training data in training (60%) and testing (40%) data.

```{r datasplitting, echo=TRUE}
inTrain <- createDataPartition(org_training_data$classe, p=0.6, list=FALSE)
training <- org_training_data[inTrain,]
testing <- org_training_data[-inTrain,]
dim(training)
dim(testing)
```

## Decision Tree Model
```{r decisiontreemodel, echo=TRUE}
DT_modfit <- train(classe ~ ., data = training, method="rpart")
DT_prediction <- predict(DT_modfit, testing)
confusionMatrix(as.factor(DT_prediction), as.factor(testing$classe))
rpart.plot(DT_modfit$finalModel, roundint=FALSE)
```
We can see that the prediction accuracy is 50% which is not upto the desired level.


## Random Forest Model

```{r randomforestmodel, echo=TRUE}
RF_modfit <- train(classe ~ ., data = training, method = "rf", ntree = 100)
RF_prediction <- predict(RF_modfit, testing)
RF_pred_conf <- confusionMatrix(as.factor(RF_prediction), as.factor(testing$classe))
RF_pred_conf
plot(RF_pred_conf$table, col = RF_pred_conf$byClass, 
     main = paste("Random Forest - Accuracy Level =",
                  round(RF_pred_conf$overall['Accuracy'], 4)))
```

##Gradient Boosting Model

```{r gradientboostingmodel, echo=TRUE}
GBM_modfit <- train(classe ~ ., data = training, method = "gbm", verbose = FALSE)
GBM_modfit$finalModel
GBM_prediction <- predict(GBM_modfit, testing)

GBM_pred_conf <- confusionMatrix(as.factor(GBM_prediction), as.factor(testing$classe))
GBM_pred_conf
plot(GBM_pred_conf$table, col = GBM_pred_conf$byClass, 
     main = paste("Gradient Boosting - Accuracy Level =",
                  round(GBM_pred_conf$overall['Accuracy'], 4)))
```

Now we need to see how each model has predicted the validation dataset across the classifications. We are not considering Decision Tree model as it didn’t reach the satisfactory prediction accuracy level. SO only Random Forest and Gradient Boosting methods are being compared.

```{r compare, echo=TRUE}
RF_pred_conf$overall

GBM_pred_conf$overall
```


## Conclusion
the Random Forest model has definitely more accuracy than GBM. Hence we will be selecting Random Forest model for final prediction from org_testing_data .


### Expected out-of-sample error
The expected out-of-sample error is estimated at 0.005, or 0.5%. The expected out-of-sample error is calculated as 1 - accuracy for predictions made against the cross-validation set. Our Test data set comprises 20 cases. With an accuracy above 99% on our cross-validation data, we can expect that very few, or none, of the test samples will be missclassified.

### Final Prediction- Applying selected model on the Test Data
```{r finalprediction, echo=TRUE}
Final_RF_prediction <- predict(RF_modfit, org_testing_data )
Final_RF_prediction
```

